### 2023-09-29 manuscript.tex
Transferred some changes from the proofs

\textbf{Corresponding author email address:}\\
* marcsingleton@berkeley.edu

@@

\section{Abstract}
Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology. However, methods for inferring and aligning such sequences in annotated genomes have not kept pace with the increasing scale and complexity of the available data. Thus, in this work we implemented several improvements to the traditional methodology that more fully leverage the redundancy of closely related genomes and the organization of their annotations. Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic HMM. In making the latter, we wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience. We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,566 high-quality alignments.

@@

Given this importance, methods for inferring orthologous groups of proteins were developed shortly after the first genomes were sequenced in the late 1990s~\cite{Fleischmann1995, Goffeau1996, CESC1998}. One early and influential approach was to cluster triangles of hits resulting from homology searches between pairs of genomes~\cite{Tatusov1997}. Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps. Despite these improvements, many databases and pipelines use the same triangle clustering algorithm~\cite{Jensen2007} or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components and other single-linkage criteria~\cite{Remm2001, Train2017, Linard2011, Cosentino2018} or Markov clustering~\cite{Enright2002, Li2003, Emms2015}. However, the scale of biological sequence data has changed dramatically. For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig.~\ref{sfig:ncbi}). Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K~\cite{Thomas2020}, the Bird 10,000 Genomes Project~\cite{Feng2020}, and the Vertebrate Genomes Project~\cite{Rhie2021} have already yielded results. Thus, the dense taxonomic sampling made possible by these projects poses new challenges and opportunities for the standard methods of orthology inference and alignment, which implicitly assume fewer and more distantly related genomes or fail to fully leverage the redundancy and organization of their annotations.

@@

\subsection{Alignment curation}
The alignments were coded into gap and non-gap symbols to simplify the emission distributions. The ``insertion'' phylo-HMM was composed of the four hidden states described in the main text. The emission distributions for each consisted of two components which modeled the gap pattern and the propensity for those patterns to remain constant (``gap stickiness''), respectively. The first component was a two-state Markov process which was in turn composed of two subprocess. The first was a phylogenetic process on the on the GTR2 consensus tree described in the section on inferring species trees, and the second was jump process at the tips. The second component was a beta-Bernoulli distribution on the number of symbols which were constant between subsequent columns. The ``missing data'' phylo-HMM was composed of two hidden states which were both parameterized with the same two-state, two component Markov process as the insertion phylo-HMM. However, the emission probabilities were calculated as the posterior probability of the observed symbol given the data rather than the probability of the data. Only the alignments of the single copy orthologous groups were curated, so each tip in the species tree corresponded to a single sequence in the alignment.

@@

\subsection{Input genomes and pre-processing}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations generally have highly complete gene sets as measured by BUSCO scores~\cite{Thibaud-Nissen2013, Simo2015}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}. In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}.

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several hidden states~\cite{Felsenstein1996, Siepel2004}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified three distinct types of regions in the alignments, roughly corresponding to regions with strong phylogenetic evidence, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1, 2, and 3, respectively. To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left). The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by noise from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right). The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocesses roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns. Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.

@@

Excluding the placement of \textit{D. obscura} (dobs), both trees have an identical topology, which is consistent with other published phylogenies. Additionally, three other nodes have short internal branches with bootstrap percentages below 100. These results indicate that not all sites and inference methods support the same topology, potentially as a result of incomplete lineage sorting. In such cases, gene trees can differ from the species tree, and we caution against using the species tree in gene-level analyses without further investigation if deviations from the true gene trees can significantly impact the results. However, as the internal branches are generally long and the bootstrap percentages for nearly all nodes are 100, the tree is generally highly resolved and therefore appropriate for use in population- or genome-level analyses without further incorporation of uncertainty.

>>>>

\textbf{Corresponding author:}\\
* Howard Hughes Medical Institute, UC Berkeley, Berkeley, CA 94720, USA. Email: marcsingleton@berkeley.edu

@@

\section{Abstract}
Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology. However, methods for inferring and aligning such sequences in annotated genomes have not kept pace with the increasing scale and complexity of the available data. Thus, in this work we implemented several improvements to the traditional methodology that more fully leverage the redundancy of closely related genomes and the organization of their annotations. Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic hidden Markov model (HMM). In making the latter, we wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience. We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,566 high-quality alignments.

@@

Given this importance, methods for inferring orthologous groups of proteins were developed shortly after the first genomes were sequenced in the late 1990s~\cite{Fleischmann1995, Goffeau1996, CESC1998}. One early and influential approach was to cluster triangles of hits resulting from homology searches between pairs of genomes~\cite{Tatusov1997}. Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps. Despite these improvements, many databases and pipelines use the same triangle clustering algorithm~\cite{Jensen2007} or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components and other single-linkage criteria~\cite{Remm2001, Train2017, Linard2011, Cosentino2018} or Markov clustering~\cite{Enright2002, Li2003, Emms2015}. However, the scale of biological sequence data has changed dramatically. For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900~\cite{OLeary2016} (Fig.~\ref{sfig:ncbi}). Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K~\cite{Thomas2020}, the Bird 10,000 Genomes Project~\cite{Feng2020}, and the Vertebrate Genomes Project~\cite{Rhie2021} have already yielded results. Thus, the dense taxonomic sampling made possible by these projects poses new challenges and opportunities for the standard methods of orthology inference and alignment, which implicitly assume fewer and more distantly related genomes or fail to fully leverage the redundancy and organization of their annotations.

@@

\subsection{Alignment curation}
The alignments were coded into gap and non-gap symbols to simplify the emission distributions. The ``insertion'' phylogenetic hidden Markov model (phylo-HMM) was composed of the four hidden states described in the main text. The emission distributions for each consisted of two components which modeled the gap pattern and the propensity for those patterns to remain constant (``gap stickiness''), respectively. The first component was a two-state Markov process which was in turn composed of two subprocess. The first was a phylogenetic process on the on the GTR2 consensus tree described in the section on inferring species trees, and the second was jump process at the tips. The second component was a beta-Bernoulli distribution on the number of symbols which were constant between subsequent columns. The ``missing data'' phylo-HMM was composed of two hidden states which were both parameterized with the same two-state, two component Markov process as the insertion phylo-HMM. However, the emission probabilities were calculated as the posterior probability of the observed symbol given the data rather than the probability of the data. Only the alignments of the single copy orthologous groups were curated, so each tip in the species tree corresponded to a single sequence in the alignment.

@@

\subsection{Input genomes and pre-processing}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database~\cite{OLeary2016}. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations generally have highly complete gene sets as measured by BUSCO scores~\cite{Thibaud-Nissen2013, Simo2015}. The \textit{Drosophila miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{Drosophila melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}. In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}.

@@

Phylogenetic hidden Markov models (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several hidden states~\cite{Felsenstein1996, Siepel2004}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified three distinct types of regions in the alignments, roughly corresponding to regions with strong phylogenetic evidence, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1, 2, and 3, respectively. To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left). The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by noise from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right). The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocesses roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns. Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.

@@

Excluding the placement of \textit{Drosophila obscura} (dobs), both trees have an identical topology, which is consistent with other published phylogenies~\cite{D12GC2007, DaLage2007}. Additionally, three other nodes have short internal branches with bootstrap percentages below 100. These results indicate that not all sites and inference methods support the same topology, potentially as a result of incomplete lineage sorting. In such cases, gene trees can differ from the species tree, and we caution against using the species tree in gene-level analyses without further investigation if deviations from the true gene trees can significantly impact the results. However, as the internal branches are generally long and the bootstrap percentages for nearly all nodes are 100, the tree is generally highly resolved and therefore appropriate for use in population- or genome-level analyses without further incorporation of uncertainty.
